---
title: "Creating D3 Chord Diagrams"
author: "Matthias Flor"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
        toc: yes
        fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Creating D3 Chord Diagrams}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

The `chorddiag` package allows to create interactive chord diagrams using the JavaScript visualization library D3 (http://d3js.org) from within R via the `htmlwidgets` interfacing framework.

In short, chord diagrams show directed relationships among a group of entities.
The chord diagram layout is explained by Mike Bostock, the creator of D3, in more detail here: https://github.com/mbostock/d3/wiki/Chord-Layout.

To quote the explanation found there:

> "Consider a hypothetical population of people with different hair colors: black, blonde, brown and red. Each person in this population has a preferred hair color for a dating partner; of the 29,630 (hypothetical) people with black hair, 40% (11,975) prefer partners with the same hair color. This preference is asymmetric: for example, only 10% of people with blonde hair prefer black hair, while 20% of people with black hair prefer blonde hair. A chord diagram visualizes these relationships by drawing quadratic Bézier curves between arcs. The source and target arcs represents two mirrored subsets of the total population, such as the number of people with black hair that prefer blonde hair, and the number of people with blonde hair that prefer black hair."
([Mike Bostock](https://github.com/mbostock/d3/wiki/Chord-Layout))

The package's JavaScript code is based on http://bl.ocks.org/mbostock/4062006, with modifications for fading behaviour and addition of tooltips.


# Basic Usage: Hair Color Preference Example

To create a chord diagram for the hair color preference example stated in the intoduction, we need the preferences in matrix format:

```{r}
m <- matrix(c(11975,  5871, 8916, 2868,
              1951, 10048, 2060, 6171,
              8010, 16145, 8090, 8045,
              1013,   990,  940, 6907),
            byrow = TRUE,
            nrow = 4, ncol = 4)
haircolors <- c("black", "blonde", "brown", "red")
dimnames(m) <- list(have = haircolors,
                    prefer = haircolors)
```

So, row names show what hair color people have, and column names show what hair color they prefer:

```{r, echo = FALSE, results = 'asis'}
pander::pandoc.table(m)
```

Then, we can pass this matrix to the `chorddiag` function to create the chord diagram:

```{r, fig.width = 8}
groupColors <- c("#000000", "#FFDD89", "#957244", "#F26223")
chorddiag::chorddiag(m, groupColors = groupColors, groupnamePadding = 20)
```


This diagram will be interactive when generated by the function.
Interactive here means chord fading and tooltip popups on certain mouse over events.
E.g. if the mouse pointer hovers over the chord connecting the "blonde" and "red" groups, a tooltip is displayed giving the numbers for the chord, and all other chords fade away.

The default chord diagram type is **directional** but chord diagrams can also be a useful visualization of relationships between two categories of groups, i.e. contingency tables.
In this package, this type of chord diagram is called **bipartite** (because there are only chords *between* categories but not *within* categories).
Here is an example for the `Titanic` dataset, depicting how many passengers from the different classes and from the crew survived or died when the ship sunk:

